---
title: "kokoszka FDA CH1"
author: Noa Jeong 
output: 
  
  pdf_document : default
  html_document : default
geometry: margin=1in
date: "2024-07-12"
---


- github : 
https://minitistics.tistory.com/27

# 1. First steps in the analysis of functional data 
```{r}
rm(list=ls())
```

**You need to update R version (current: R 4.4.1)** : update R (not R studio)
```{r}
#install.packages("fda")
library(fda)
```


$$
x_n(t_{j,n}) \in \mathbb{R} \quad t_{j,n} \in [T_1, T_2] \quad n=1,2, ..., N \quad j = 1,...,J_n
$$

$N$ : number of curves
$J_n$ : observed time point 

\newpage
## 1.1 Basis Expansion 

$$
X_n(t) \approx \sum_{m=1}^N c_{nm} B_{m}(t) 
$$

$B_m$ : basis functions (splines wavelets, sine cosine functions)


- 5 B-spline basis functions defined on the interval [0,10]. 

```{r}
spline.basis = create.bspline.basis(rangeval=c(0,10), # interval
                                    nbasis=5) # number of basis functions 
plot(spline.basis,lty=1,lwd=2)
```

- first five Fourier basis functions 
```{r}
fourier.basis = create.fourier.basis(rangeval=c(0,10),
                                     nbasis=5)
plot(fourier.basis)
```

Fourier system is usually only suitalbe for functions which have approximately the same values at the beginnin and the end of the interval. 

### Example 1.1.1 [B-spline expansion on Wiener process]

appropriate random walk 
$$
S_i = \frac{1}{\sqrt K} ,\quad N_k \sim iid \, N(0,1)
$$

- Random walk and its expansion using 25 B-Spline basis functions 
```{r}
Wiener = cumsum(rnorm(10000)/100) # random walk on [0,K], K=10^4
plot.ts(Wiener, xlab="", ylab="")

B25.basis = create.bspline.basis(rangeval=c(0,10000),
                                 nbasis=25)
Wiener.fd = smooth.basis(y=Wiener, fdParobj=B25.basis) # create functional data object Wiener.fd 
lines(Wiener.fd, lwd=3)
```



\newpage 

## 1.2 Sample Mean and covariance 
raw data -> functional objects. by suitable basis expansion 


**pointwise mean & pointwise standard deviation** 
$$
\bar{X}_N(t) = \frac{1}{N} \sum_{n=1}^N X_n(t) \quad SD_X(t) = \{\frac{1}{N-1} \sum_{n=1}^N (X_n(t)-\bar{X}_N(t))^2 \} ^{1/2}
$$

### Example 1.2.1 [Pointwise mean and SD]
```{r}
N=50
W.mat=matrix(0, ncol=N, nrow=10000) 
for(n in 1:N){
  W.mat[,n] = cumsum(rnorm(10000))/100
}

B25.basis = create.bspline.basis(rangeval=c(0,10000),
                                 nbasis=25)
W.fd = smooth.basis(y=W.mat,
                    fdParobj = B25.basis)

plot(W.fd,ylab='',xlab='',col='gray',lty=1)
W.mean <- mean.fd(W.fd$fd)
W.sd <- std.fd(W.fd$fd)
lines(W.sd, lwd=3);lines(W.mean,lty=2,lwd=3)
```
pointwise s.d : typical variablity of curves at time point *t* 

But no information on how values of curve at point *t* relate to point *s*


- **sample covariance function**
$$
\hat c(t,s) = \frac{1}{N-1}\sum_{n=1}^N (X_n(t)-\bar X_N(t)) (X_n(s)-\bar X_N(s))
$$
interpretaion as variance covariance matrix 

### Example 1.2.2 [Sample covariance function]

```{r}
# Use the object W.fd generated in the previous example.
W.cov = var.fd(W.fd$fd) # $fd extracts function values
grid=(1:100)*100

W.cov.mat=eval.bifd(grid,grid,W.cov)
persp(grid,grid,W.cov.mat,xlab='s',
      ylab='t',zlab='c(s,t)')
contour(grid,grid,W.cov.mat,lwd=2)
```

$\hat c(t,s)$ is given by $c(t,s)=min(t,s)$


\newpage 

## 1.3 Principal component functions 

EFPC's : Estimated fuctional principal components     
related to the sample covariance function $\hat c(t,s)$


- ceneterd function 
$$
X_n(t) - \bar X_n(t) \approx \sum_{j=1}^p \ \xi_{n j} \hat v_j(t)
$$

p is much smaller than M

$\hat v_j(t)$ are computed from the observed functions $X_1, X_2, .. X_N$ after converting them to functional objects 



```{r}
W.pca = pca.fd(W.fd$fd, nharm=4)
plot(W.pca$harmonics, lwd=3)
W.pca$varprop
```

$\hat v_1$ black line : most pronounced pattern of the deviation from the mean function of a randomly selected trajectory. 

coefficient $\xi _{n1}$ quantifies the contribution of $\hat v_1$ to its shape. 

$\hat v_2$ red line : second most important mode of mean functions of 50 random walks. It is second msot important mode of variability which is orthogonal to $\hat v_1$



$\xi _{nj}$ is called the score of $X_n$ with respect to $\hat v_j$ .; the smaller the percentages, the smaller the scores. 

EFPC's $\hat v_j$ are orthonormal 
$$
\int \hat v_j(t) \hat v_i(t) dt =  
\left\{ 
  \begin{array}{ c l }
    0 & \quad \textrm{if } j \ne i \\
    1 & \quad \textrm{if } j = i
  \end{array}
\right.
$$

\newpage 

## 1.4 Analysis of BOA stock returns 

Bank of America 

1997.4.9 - 2007.4.2, 9:30 ~ 16:00 stock values recorded every minute 

$t \in (0,6.5)$

2511 days of data, each day consisting 390 measurements   



functional observation : *cumulative log-return*
$$
R_n(t) := log(P_n(t)) - log(P_n(0)) \approx \frac{P_n(t)-P_n(0)}{P_n(0)}
$$

$R_n(t)$ : how an investment, made at opening, evolves over the course of the day 

$P_n(t)$ : value of the stock on the day n at time t

- plot of first ten cumulative log returns for BOA

```
#install.packages('striprtf')
#install.packages('readr')
library(striprtf)
library(readr)

extracted_text <- striprtf::read_rtf('Bank of America Data.rtf')
```


```{r}
BOA = read.table('BOA.txt', header=TRUE)
Dates <- dimnames(BOA)[[1]] # 2511 days
BOA <- data.matrix(BOA)
Outlier = which(Dates=='08/26/2004')
BOA <- BOA[-Outlier,]
```

- Plot of first 10 cumulative log returns for BOA 
```{r}
N <- dim(BOA)[1] # 2510 
M <- dim(BOA)[2] # 390 measurements 
Times <- seq(0,6.5,length=M)
log_BOA <- log(BOA)-matrix(log(BOA)[,1], nrow=N, ncol=M) # R_n(t)
bspline_basis <- create.bspline.basis(rangeval=c(0,6.5), norder=4, nbasis=200)
log_BOA_f <- Data2fd(Times,t(log_BOA), basisobj = bspline_basis)
plot(log_BOA_f[1:10], xlab='',ylab='',lwd=1.5)
```

- plot of mean function of BOA cumulative returns. Pointwise 95% confidence intervals are included in red. 
```{r}
muhat <- mean.fd(log_BOA_f)
sdhat <- sd.fd(log_BOA_f)
SE_hat_U <- fd(basisobj = bspline_basis) # create upper CI bound 
SE_hat_L <- fd(basisobj = bspline_basis) # create lower CI bound
SE_hat_U$coefs <- 2*sdhat$coefs/sqrt(N) + muhat$coefs
SE_hat_L$coefs <- -2*sdhat$coefs/sqrt(N) + muhat$coefs
plot.fd(SE_hat_U, ylim=c(-0.002,0.002), col='red',lty=2,xlab='',ylab='')
plot.fd(SE_hat_L, add=T, col='red', lty=2)
plot.fd(muhat, add=T)
```


- plot of first four PCs of BOA 
```{r}
log_BOA_pca <- pca.fd(log_BOA_f,nharm=4)
plot(log_BOA_pca$harmonics,lwd=1)
```




\newpage 
## 1.5 Diffusion tensor imaging  
DTI : magnetic resonance imaging methodology, to measure the diffusion of water in the brain.   Utilize DTI to generate image of white matter in the brain. 

Fractional anisotropy is a value 0~1 which measure the level of anisotropy, therefore the quantity of white matter at a particular location. 

376 patients, each tract measure at 93 equally spaced location. 

$t_{j,n}$ : spatial rather than temporal location. 
```{r}
#install.packages('refund')
library(refund)
data(DTI)
str(DTI)
```


```{r}
Y <- DTI$cca
Y <- Y[-c(126,130,131,125,319,321),] # missing values
N <- dim(Y)[1]; M <- dim(Y)[2] 
# N : number of patients(observations/functions)
# M : number of locations(time/spatial observed point)

argvals <- seq(0,1,length=M)
data_basis <- create.bspline.basis(c(0,1), # interval
                                   nbasis=10) # number of basis
Y.f <- Data2fd(argvals, t(Y), data_basis) # create smooth functions that fit scatter plot data 

```


- plot of fractional anistoropy tract profiles oft he corpus collosum. 
```{r}
#dev.new(width=8,height=6)
plot(Y.f,lty=1,col='gray',xlab='',ylab='',ylim=c(0.1,0.9)) # Individuals(376 patients) tracts are plotted 
lines(mean.fd(Y.f), lwd=2) # mean function 
# green, yellow, red indicates respecitvely one, two, three pointwise s.d. from the mean. 
lines(mean.fd(Y.f)+std.fd(Y.f), lwd=2,lty=2,col='green')
lines(mean.fd(Y.f)-std.fd(Y.f), lwd=2,lty=2,col='green')
lines(mean.fd(Y.f)+2*std.fd(Y.f), lwd=2,lty=2,col='yellow')
lines(mean.fd(Y.f)-2*std.fd(Y.f), lwd=2,lty=2,col='yellow')
lines(mean.fd(Y.f)+3*std.fd(Y.f), lwd=2,lty=2,col='red')
lines(mean.fd(Y.f)-3*std.fd(Y.f), lwd=2,lty=2,col='red')
```

data follows an analog of the classic three standard deviation rule well, with nearly all curves falling between the red lines 






\newpage 
## 1.6 Chapter 1 Problems 

### 1.1 
```{r}
rm(list=ls())
library(fda)
data(pinch)

str(pinch) # 151 measurements of pinch for 20 replications(curves)
```


#### (a)

Convert the pinch data to functional objects using B-splines of order four(cubic splines) and plot the 20 smoothed curves on one graph 
```{r}
dim(pinch)
N <- dim(pinch)[1]; M <- dim(pinch)[2] # 151 20
```




```{r}
argvals <- seq(1,20) # 20 replications 
spline.basis <- create.bspline.basis(rangeval = c(1,20),
                                     nbasis=15, norder=4)

pinch.f <- Data2fd(argvals,t(pinch), spline.basis ) 
plot(pinch.f, col='gray',xlab='',ylab='')
```

#### (b)

Calculate the pointwise mean and SD and add them to plot. 

```{r}
mean.pinch <- mean.fd(pinch.f)
sd.pinch <- sd.fd(pinch.f)
plot(pinch.f, col='gray',xlab='',ylab='',ylim=c(-5,11))
lines(mean.pinch, lwd=2)
lines(mean.pinch+sd.pinch, col='red',lwd=2,lty=2)
lines(mean.pinch-sd.pinch, col='red',lwd=2,lty=2)
lines(mean.pinch+2*sd.pinch, col='blue',lwd=2,lty=2)
lines(mean.pinch-2*sd.pinch, col='blue',lwd=2,lty=2)
```

#### (c)
Graph the persepctive and contour plots of the sample covariance function $\hat c(t,s)$ of the pinch curves. 
```{r}
#pinch.f$fd
cov.pinch <- var.fd(pinch.f)
#cov.pinch$coefs
grid <- seq(1,20,length=N)

cov.mat.pinch <- eval.bifd(grid,grid,cov.pinch)
persp(grid,grid,cov.mat.pinch,
      xlab='s',ylab='t',zlab='c(t,s)')
contour(grid,grid,cov.mat.pinch)
```


#### (d)

Graph the first four EFPC's of the pinch data. How many components do you need to explain 90% of variation? 

```{r}
pinch.pca <- pca.fd(pinch.f, nharm=4)
plot(pinch.pca$harmonics, lwd=3)

pinch.pca$varprop # only one component needed to explain 90% of variation 
```


\newpage

### 1.2 
United states Federal Reserve interest rates `FedYiedlcurve`, which contains the monthly interest rates from January 1982 to June 2009 

x values : maturity terms of 3,6,12,60,84,120 months. $t_j$ 

y values : interest rates of the U.S. Treasury obligations due in x months. $x_n(t_j)$ , where $n$ is a month in. the range 1982.1 to 2009.6 

```{r}
rm(list=ls())
library(fds);library(fda)
```

#### (a) 
On one graph, plot the interest rate $x(t_j)$ for Jan 1982 - Jun 2009 against the maturity terms $t_j$. How do the interest rates in these 2 months compare? 
```{r}
data(FedYieldcurve)
yield <- FedYieldcurve ; terms = yield$x
plot(terms, yield$y[,1], pch=15, ylab='Yield', ylim=c(0,16)) # square 
points(terms, yield$y[,330], pch=16) # circle . lower interest rates at June 2009
```

#### (b)   

Convert the yield data to functional objects using bspline basis with 4 basis functions. Calculate and plot the mean yield function. What is the average behavior of interest rates as function of maturity? 

```{r}
max(yield$y)
min(yield$y)
M <- dim(yield$y)[1] ; N <- dim(yield$y)[2] # 6 330

bspline_basis <- create.bspline.basis(range=c(3,120), nbasis=4)
yield.fd <- Data2fd(terms,yield$y,basisobj = bspline_basis)

plot(yield.fd, lty=1, col='gray', xlab='terms', ylab='yield')

yield.mean <- mean.fd(yield.fd)
lines(yield.mean,lwd=2,col='red')
```

#### (c)
```{r}
yield.pca <- pca.fd(yield.fd, nharm=1)
plot(yield.pca$harmonics,lwd=3)

yield.pca$varprop
```
first component explains a lot of patterns of the deviation from the mean functions of yield. 

\newpage 
### 1.3 
Data for monthly SST(in degrees C) in various regions of the south Pacific ocean from 1950 to 2013. Using the `NINO3` column, treat data for each calander year as a functional observation. 
```{r}
rm(list=ls())
?sd.fd
```

```{r}
nino <- read.table('ninoSST.txt')
head(nino)

library(fda)
year <- nino[,1]
month <- 1:12

#SSTmat <- matrix(data=NA, nrow=12, ncol=64)
#count=1
#for(yr in year){
#  tmp <- nino[which(nino$V1==yr),]$V4
#  SSTmat[,count] <- tmp
#  count <- count+1
#}

SSTmat <- as.matrix(t(nino[,-1]))
rownames(SSTmat) = NULL
colnames(SSTmat) <- seq(1950,2013)
dim(SSTmat) # month, 1950-2013 years

head(SSTmat)
```

#### (a) 
Using Fourier basis functions, convert the data into a functional object `ninofd` containing 64 annuals curves and plot these smoothed curves. 
```{r}
N <- dim(SSTmat)[2] # 64
M <- dim(SSTmat)[1] # 12 
argvals <- seq(1,12)
fourier.basis = create.fourier.basis(rangeval=c(1,12),
                                     nbasis=M)
ninofd <- Data2fd(argvals, SSTmat, fourier.basis)
plot(ninofd, col='gray')
lines(mean.fd(ninofd),col='black',lwd=3)
```

#### (b)
The years of 1965, 1972, 1982, 1997 were pronounced El Nino years. emphasize the curves in the plot by making them thick red. 

Emphasize the pronounced La Nina years of 1955, 1973, 1975, 1988, 1999 with thick blue. 
```{r}
#Pronounced ElNino Years - characterized by unusually warm SST
eln = c(1965, 1972, 1982, 1997)
#Pronounced La Nina Years - characterized by unusually cold SST
lan = c(1955, 1973, 1975, 1988, 1999)

elnn = match(eln,year)
lann = match(lan,year)
ELNmat = SSTmat[,elnn]
LANmat = SSTmat[,lann]

ELNfd <- Data2fd(argvals, ELNmat, fourier.basis)
LANfd <- Data2fd(argvals, LANmat, fourier.basis)
plot(ninofd, col='gray')
lines(mean.fd(ninofd),col='black',lwd=3)
lines(ELNfd,col='red',lwd=3,lty=1)
lines(LANfd,col='blue',lwd=3,lty=1)
```

\newpage 
### 1.4 

\newpage 
### 1.5 
Matern covariance function leads to general family of stationary Gaussian processes. 

- covariance function 
$$
C(t,s) = \frac{\sigma^2}{\Gamma(\nu) 2^{\nu-1}} (\frac{\sqrt{2\nu} |t-s|}{\rho})^{\nu} K_{\nu}(\frac{\sqrt{2\nu} |t-s|}{\rho}), \quad \nu>0
$$

$\sigma^2$ : variance parameter     
$\nu$ : smoothness parameter.    
$\rho$ : range parameter     
$K_{\nu}$ : modified Bessel function of the second kind.     

k times continuously differentiable for any $\nu$.   




#### (1)

Simulate plot iid mean zero Marten process with 
$\nu = 1/2, \, \nu = 2, \, \nu = 4$. Set $\sigma^2=1$ and $\rho=1$. 
temporal grid with 50 evenly spaced points. 

#### (2) 
Plot the first four EFPCS for each value of $\nu$, comment on any similarities / differences 

#### (3)
Plot the explained variance for each $\nu$, comment on any similarities/differences.

#### (4)
Using your preferred method, create a plot of the covariance surface for each $\nu$, comment on any similarities/differences.



```{r}
matern_cov <- function(t,s,sigma2,nu,rho){
  if(t==s){
    return(sigma2)
  } else{
  factor1 <- sigma2 / (gamma(nu) * 2^(nu-1))
  factor2 <- (sqrt(2*nu)*abs(t-s) / rho)^nu
  factor3 <- besselK(sqrt(2*nu)*abs(t-s)/rho, nu)
  return(factor1*factor2*factor3)
  }
} # each element in covariance 

# parameters 
sigma2 <- 1
rho <- 1
nu_values <- c(0.5,2,4)
time_points <- seq(0, 1, length.out = 50)

# create 50x50 matern_cov matrix 
create_cov_mat <- function(time_points, sigma2, nu, rho){
  n <- length(time_points)
  cov_mat <- matrix(0,n,n)
  for(i in 1:n){
    for(j in 1:n){
      cov_mat[i,j] <- matern_cov(time_points[i], time_points[j],sigma2, nu, rho)
    }
  }
  return(cov_mat)
}
  
# simulation functions 
simul <- function(n_sim, time_points, sigma2, nu, rho){
  n <- length(time_points)
  cov_mat <- create_cov_mat(time_points, sigma2, nu, rho)
  simulation <- mvrnorm(n_sim, mu=rep(0,n), Sigma = cov_mat)
  return(simulation)
}
```

```{r fig.align='center'}
set.seed(1)
# Simulation 100 times 

for(nu in nu_values){
  simul_result <- simul(100, time_points, sigma2,nu,rho) 
  # (1)
  # fda package 
  matplot(time_points,t(simul_result),type='l',col='gray',lty=1,
          xlab='Time',ylab='Value',main=paste('nu =',nu))
  
  # (2)
  basis <- create.bspline.basis(rangeval=c(0,1), nbasis=5)
  fd_obj <- Data2fd(time_points,t(simul_result), basis)
  pca_result <- pca.fd(fd_obj,nharm=4)
  plot(pca_result$harmonics,lwd=3)
  
  # (3)
  explained_variance <- cumsum(pca_result$values)/sum(pca_result$values)*100
  plot(explained_variance,type='b')
  
  
  # (4)
  fd_cov <- var.fd(fd_obj)
  grid <- seq(0,1,length=50)
  fd_cov_mat <- eval.bifd(grid,grid,fd_cov)
  persp(grid,grid,fd_cov_mat,xlab='s',ylab='t',zlab='C(t,s)')
  contour(grid,grid,fd_cov_mat)

}
```





\newpage (GPT code..)
```{r}
# Load necessary library for matrix operations
library(MASS)

# Define the covariance function for the Matern process
matern_cov <- function(h, nu, sigma2, rho) {
  if (h == 0) {
    return(sigma2)
  } else {
    factor1 <- (2^(1 - nu)) / gamma(nu)
    factor2 <- (sqrt(2 * nu) * h / rho) ^ nu
    factor3 <- besselK(sqrt(2 * nu) * h / rho, nu)
    return(sigma2 * factor1 * factor2 * factor3)
  }
}

# Parameters
sigma2 <- 1
rho <- 1
nu_values <- c(0.5, 2, 4)
time_points <- seq(0, 1, length.out = 50)

# Function to create covariance matrix
create_cov_matrix <- function(time_points, nu, sigma2, rho) {
  n <- length(time_points)
  cov_matrix <- matrix(0, n, n)
  for (i in 1:n) {
    for (j in 1:n) {
      cov_matrix[i, j] <- matern_cov(abs(time_points[i] - time_points[j]), nu, sigma2, rho)
    }
  }
  return(cov_matrix)
}

# Simulation function
simulate_matern_process <- function(n_sim, time_points, nu, sigma2, rho) {
  n <- length(time_points)
  cov_matrix <- create_cov_matrix(time_points, nu, sigma2, rho)
  simulations <- mvrnorm(n_sim, mu = rep(0, n), Sigma = cov_matrix)
  return(simulations)
}

# Set seed for reproducibility
set.seed(123)

# Plotting function
plot_simulations <- function(simulations, time_points, nu) {
  matplot(time_points, t(simulations), type = 'l', col = 'gray', lty = 1, 
          main = paste('Matérn Process Simulations (nu =', nu, ')'), 
          xlab = 'Time', ylab = 'Value')
  grid()
}

# Simulate and plot for each nu value
par(mfrow = c(3, 1)) # 3 plots in one column
for (nu in nu_values) {
  simulations <- simulate_matern_process(100, time_points, nu, sigma2, rho)
  plot_simulations(simulations, time_points, nu)
}

```



