---
title: "kokoszka FDA CH12 "
author: Noa Jeong 
output: 
  pdf_document:  
    latex_engine: xelatex
  html_document: default
---
# Ch 12 

```{r message=FALSE}
rm(list=ls())

library(MASS)
library(fda)
library(ggplot2)
```


## 12.4

```{r}
# R.version # R version 4.4.1 (2024-06-14)
```

```{r message=FALSE}
#install.packages("sp")
#install.packages("RandomFieldsUtils_1.2.5.tar.gz", repos = NULL)
#install.packages('RandomFields_3.3.14.tar.gz',repos = NULL)

#install.packages('CompQuadForm')

library(RandomFields); library(CompQuadForm)
```

- `RandomFields` is used to simulate the functions.     
- `CompQuadForm` is used to find the p-values for the norm approach.

### A simulation Study 

N = 100 with 50 evenly spaced time points over [0,1] 

$$
X_n(t) = \mu(t) + \epsilon_n(t)
$$

with $\epsilon_n$ taken to be iid Matern processes with variance=1, scale = 1/4, smoothness = 5/2 

mean function $\mu$ to be 
$$
\mu(t) = c_1 \sqrt2 sin((k-1/2)\pi t)
$$



First we set an option for RandomFields so that matrices are output.
```{r}
RFoptions(spConform=FALSE) # makes output a matrix
```

Then we set the parameters of the Matern error
```{r}
nu<-3/2; var = 1; scale = 1/4
Mat_model<-RMmatern(nu = nu , var = 1 , scale = scale)
N<-10; m<-50;
```

Next we generate the mean function
```{r}
times<-seq(0,1,length=m)
c1<-0; k<-1
mu<-function(x){c1*sqrt(2)*sin((k-1/2)*pi*x)}
mu_vec<-mu(times)
```

Now we simulate the data and tests.
```{r}
reps<-10
TPC3_pval<-numeric(reps)
Tnorm_pval<-numeric(reps)
for(i in 1:reps){
  # Step 1 - Simulate Data
  # Prevent RFsimulate from printing
  capture.output(eps<-RFsimulate(Mat_model,
                                 times,
                                 n=N))
  X_mat<-scale(t(eps),center=-mu_vec,scale=FALSE)
  X.f<-Data2fd(times,t(X_mat))
  # Step 2 - Estimate parameters
  muhat<-mean.fd(X.f)
  X.pca<-pca.fd(X.f,nharm=20)
  lambda<-X.pca$values
  scores<-X.pca$scores
  v<-X.pca$harmonics
  # Step 3 - Compute tests and p-values
  # PCA test with 3 PCs 
  TPC3 <- N*sum(inprod(v[1:3],muhat)^2/lambda[1:3])
  TPC3_pval[i]<-pchisq(TPC3,3,lower.tail=FALSE)
  # Norm test
  Tnorm<-N*sum(inprod(v[lambda[1:20]>0],muhat)^2)
  Tnorm_pval[i]<-imhof(Tnorm,lambda[lambda[1:20]>0])[[1]]}
```

`RFsimulate` : Simualtion of Random Fields 
simulate **unconditional** random fields 


\newpage

## 12.8 Chapter 12 Problems 

### Q. 12.3 

Recall the simulation scheme given in **Problem 1.5.** For the three $\nu$ scenarios, increase N to 1000. 


```{r message=FALSE}
rm(list=ls())
library(fda)
library(ggplot2)
#install.packages('tidyverse')
library(tidyverse)
```

(a) Plot the estimates for the first 10 eigenvalues and include confidence intervals. Include everything in one plot if you can (but use no more than three plots). 

(`?RMmatern`)

- `RMmatern` : stationary isotropic covariance model belonging to matern family.   

- `RFsimulate` : Simulate functional data based on specific covariance function 


(b) How many of the CI’s did not include 0? Suppose you used that to determine the number of FPC’s; how does that compare to the explained variance approach?

```{r}
RFoptions(spConform=FALSE)
```

```{r}
set.seed(1)

# set parameters 
nu_values <- c(0.5,2,4); sigma2 <- 1; rho <- 1 
N <- 1000 
m <- 50 # time points 

# generate mean function 
time_points <- seq(0, 1, length.out = m)
c1 <- 0; k <- 1
mu <- function(x){c1*sqrt(2)*sin((k-1/2)*pi*x)}
mu_vec <- mu(time_points) 


n_obs <- 100
n_simul <- 100
lambda_boot <- matrix(NA, nrow=n_simul, ncol=10)

# simualte data 
for(i in 1:3){ # for each nu values 
  # epsilon_n taken to b iid Matern process 
  Mat_model <- RMmatern(nu_values[i], 
                      var = sigma2, 
                      scale = rho)
  capture.output(eps <- RFsimulate(Mat_model, # cov function
                                   time_points,
                                   n=N)) # number of function to simulate 
  X_mat <- scale(t(eps), center=-mu_vec, scale=F) 
  
  for(n in 1:n_simul){
    idx <- sample(1:1000, 
                  size=n_obs, 
                  replace=T) # bootstrap  
    boot_data <- X_mat[idx, ]
    X.f_boot <- Data2fd(time_points,t(boot_data)) # n_obs x 50
    X.pca_boot <- pca.fd(X.f_boot, nharm=10) 
    lambda_boot[n,] <- X.pca_boot$values[1:10] # store lambda for each simulation 
  }
  
  lambda_mean <- apply(lambda_boot,2,mean) # 10 means 
  lambda_upper <- apply(lambda_boot, 2, quantile, probs = 0.975)
  lambda_lower <- apply(lambda_boot, 2, quantile, probs = 0.025)

  # Create data frame for plotting
  df <- data.frame(index = 1:10,
                 lambda = lambda_mean,
                 lower_ci = lambda_lower,
                 upper_ci = lambda_upper)
  print(df)

  # (a) Plot
  p <- ggplot(df, aes(x = index, y = lambda)) +
    geom_point() +
    geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci)) 
  plot(p)
  
  # (b) 
  eps <- 1e-5
  print("Number of lower bound which didn't include zero")
  print(sum(df$lower_ci > eps))

}

```




